---
title: "EV Power - Lab 4 Project Report"
format: typst
---

# Example Solution 1

## **Part 0: libraries**

```{r}

library(tidyverse)
library(maps)
```

## **Part 1:** **Defining Research Question**

Chosen Question: Electric vehicles reduce direct emissions, but does the electricity used to charge them actually come from clean sources?

**Brainstormed map questions:**
1. What percent of each state’s total energy use comes from renewables in 2023?  
2. How did renewable share change from 2021 to 2023 by state?  
3. Do states with higher renewable shares have lower average electricity prices?  
4. Are EV registrations concentrated in states with cleaner energy mixes?  
5. Which regions show the fastest growth in renewable share (2021–2023)?

**Chosen question for the map:**  
*Change in renewable energy share from 2021 to 2023 by state.*

## **Part 2: Data Preparation and Cleaning**

```{r}
library(tidyverse)

renew21 <- read_csv("data/renew-use-2021.csv")
renew22 <- read_csv("data/renew-use-2022.csv")
renew23 <- read_csv("data/renew-use-2023.csv")

total21 <- read_csv("data/total-use-2021.csv")
total22 <- read_csv("data/total-use-2022.csv")
total23 <- read_csv("data/total-use-2023.csv")

price <- read_csv("data/av-energy-price-2021-2023.csv")
evs   <- read_csv("data/ev-registrations-by-state-2023.csv")

clean_names <- function(df) {
  df %>% rename_all(~ str_to_lower(.) %>% str_replace_all("[^a-z0-9]+", "_"))
}

renew21 <- clean_names(renew21)
renew22 <- clean_names(renew22)
renew23 <- clean_names(renew23)

total21 <- clean_names(total21)
total22 <- clean_names(total22)
total23 <- clean_names(total23)

price   <- clean_names(price)
evs     <- clean_names(evs)


pivot_total <- function(df, year){
  id_col <- if ("energy_source" %in% names(df)) "energy_source" else names(df)[1]
  df %>%
    pivot_longer(
      cols = -all_of(id_col),
      names_to = "state",
      values_to = paste0("total_use_", year)
    ) %>%
    rename(energy_source = !!id_col)
}

total21 <- pivot_total(total21, 2021)
total22 <- pivot_total(total22, 2022)
total23 <- pivot_total(total23, 2023)

fix_states <- function(df){
  if (!"state" %in% names(df)) return(df)
  df %>% mutate(
    state = str_trim(state),
    state = str_to_title(state)
  )
}

renew21 <- fix_states(renew21)
renew22 <- fix_states(renew22)
renew23 <- fix_states(renew23)
total21 <- fix_states(total21)
total22 <- fix_states(total22)
total23 <- fix_states(total23)
price   <- fix_states(price)
evs     <- fix_states(evs)

make_numeric <- function(df, pattern){
  num_cols <- grep(pattern, names(df), value = TRUE)
  if (length(num_cols) > 0){
    df <- df %>%
      mutate(across(
        all_of(num_cols),
        ~ suppressWarnings(readr::parse_number(as.character(.x)))
      ))
  }
  df
}

renew21 <- make_numeric(renew21, "^renewable_use_")
renew22 <- make_numeric(renew22, "^renewable_use_")
renew23 <- make_numeric(renew23, "^renewable_use_")

total21 <- make_numeric(total21, "^total_use_")
total22 <- make_numeric(total22, "^total_use_")
total23 <- make_numeric(total23, "^total_use_")

head(renew21)
head(total21)
head(price)
head(evs)
```

## **Part 3: Joining / Pivoting Datasets for Analysis**

```{r}

library(readr)

prep_year <- function(df, value_prefix, yr) {
  vcol <- grep(paste0("^", value_prefix, "_", yr, "$"), names(df), value = TRUE)
  stopifnot(length(vcol) == 1)

  # pull the column once, coerce only if needed
  vals <- df[[vcol]]
  if (!is.numeric(vals)) {
    vals <- readr::parse_number(as.character(vals))
  }

  # ensure energy_source exists (totals should already have it; just in case)
  if (!"energy_source" %in% names(df)) {
    df$energy_source <- NA_character_
  }

  tibble::tibble(
    state         = df$state,
    year          = yr,
    energy_source = df$energy_source,
    value         = vals
  )
}

renew21_tidy <- prep_year(renew21, "renewable_use", 2021)
renew23_tidy <- prep_year(renew23, "renewable_use", 2023)

total21_tidy <- prep_year(total21, "total_use", 2021)
total23_tidy <- prep_year(total23, "total_use", 2023)

renew_by_state <- bind_rows(renew21_tidy, renew23_tidy) %>%
  group_by(state, year) %>%
  summarise(renewables = sum(value, na.rm = TRUE), .groups = "drop")

total_by_state <- bind_rows(total21_tidy, total23_tidy) %>%
  group_by(state, year) %>%
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

energy_joined <- renew_by_state %>%
  inner_join(total_by_state, by = c("state", "year")) %>%
  mutate(renew_share = (renewables / total) * 100)

renew_change <- energy_joined %>%
  select(state, year, renew_share) %>%
  tidyr::pivot_wider(
    names_from  = year,
    values_from = renew_share,
    names_prefix = "share_"
  ) %>%
  mutate(change_2021_2023 = share_2023 - share_2021) %>%
  arrange(desc(change_2021_2023))

cat("renew21 columns: ", paste(names(renew21), collapse = ", "), "\n")
cat("total21 columns: ", paste(names(total21), collapse = ", "), "\n")
head(renew_change)
```

## **Part 4: Mapping Visualization**

```{r}
stopifnot(exists("renew_change"))

us_map <- ggplot2::map_data("state") |> as_tibble()  

map_df <- renew_change |>
  mutate(region = tolower(state)) |>
  inner_join(us_map, by = "region")

ggplot(map_df, aes(x = long, y = lat, group = group, fill = change_2021_2023)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_gradient2(
    name = "Change in renewable\nshare (pp, 2023–2021)",
    midpoint = 0, low = "#b2182b", mid = "white", high = "#2166ac", na.value = "grey90"
  ) +
  labs(
    title = "Change in Renewable Energy Share by State (2021 → 2023)",
    subtitle = "Positive values indicate an increase",
    caption = "Data: renew-use-*.csv and total-use-*.csv"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

## Part 5: Final Deliverable

### 1. Title  
**Clean Energy and Electric Vehicles**  
_Project 4 — Stat 133 Lab 4 Report_



### 2. Overview  
This project explores whether the electricity that powers electric vehicles actually comes from clean sources.  
My chosen sub-question is:  
> **How did the renewable-energy share change from 2021 to 2023 by state?**  

Understanding this helps connect EV growth to the cleanliness of each state’s energy mix.


### 3. Data and Methods  
Four datasets were used:  
- `renew-use-*.csv`  (Renewable energy use by state and year)  
- `total-use-*.csv`  (Total energy use by state and year)  
- `av-energy-price-2021-2023.csv`  (Average energy price)  
- `ev-registrations-by-state-2023.csv`  (EV registrations)  

**Steps performed**
1. Cleaned and standardized column names and state labels.  
2. Joined renewable and total use datasets by (state, year).  
3. Computed `renew_share = (renewable / total) × 100`.  
4. Pivoted to compare 2021 vs 2023 and calculated `change = share_2023 − share_2021`.  

Example of joined data (`head(renew_change)`):

```{r}
head(renew_change)